Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2025-03-22 paper/paper.tex
nonlinear figs, frisch/cardelli, axioms in appendix

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{\mathsf{SPF}}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\StringSem}{\textbf{String}}
\newcommand{\CharGram}{\texttt{Char}}
\newcommand{\StringGram}{\texttt{String}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\newcommand{\ctxwff}[1]{#1 \isCtx}
\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \isTy}
\newcommand{\linctxwff}[2]{#1 \vdash #2 \isLinCtx}
\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}
\newcommand{\nonlinterm}[3]{#1 \vdash #2 : #3}
\newcommand{\linterm}[4]{#1 ; #2 \vdash #3 : #4}

@@

\section{Introduction}
\label{sec:intro}
Parsing structured data from untrusted input is a ubiquitous task in
computing. Any formally verified software system that interacts with
the outside world must contain some parsing component. For example, in
an extensive experiment finding bugs in C compilers
\cite{yangFindingUnderstandingBugs}, an early version of the formally
verified CompCert C compiler only contained bugs in the then
unverified parsing component \cite{leroy_formal_2009}. Bugs in parsers
undermine the overall correctness theorem for a verified system: an
incorrectly parsed C program will be compiled correctly but this is
not very useful if it did not correctly correspond to the actual
source program. Eventually, a correct parser was implemented using an
automaton that is formally verified to implement an LR grammar
\cite{jourdanValidatingLRParsers2012}.

@@

Prior approaches to verified parsing focus on verification of a
particular grammar formalism such as non-left-recursive grammars or
LL(1) grammars
\cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
Each new grammar formalism is extended with its own independent
verified implementation.

@@

To show the feasibility of our design, we have implemented \theoryabbv
as a shallowly embedded domain-specific language in the Cubical Agda
proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
grammars and parsers in our system including regular expressions,
non-deterministic and deterministic automata, as well as some example
context-free grammars and parsers based on LL(1) and LR(1)
automata. Throughout this paper, we will use \agdalogo~ to mark
results that are mechanized in our Agda development.

@@

The formation rules for the judgments of \theoryabbv are shown in
\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
stand for non-linear types; $M,N$ stand for non-linear terms, these
act as in an ordinary dependent type theory; $\Delta$ stands for
linear contexts; $A,B,C$ for linear types; and, $e,f,g$ for linear
terms. These contexts, types and terms are allowed to depend on an
ambient non-linear context $\Gamma$, but note that linear types $A$
cannot depend on any \emph{linear} variables in $\Delta$. We include
definitional equality judgments for both kinds of type and term
judgments as well. Additionally, we have a judgment $\Gamma\vdash X
\isSmall$ which is used in the definition of universe types.

@@

    \inferrule{\linctxwff \Gamma \Delta \and \linctxwffjdg \Gamma A}{\linctxwff
      {\Gamma} {\Delta , a : A}}

    \boxed{\ctxwffjdg \Gamma X}

    \inferrule{~}{\ctxwffjdg \Gamma {Bool}}

@@

@@

@@

    \inferrule{\lintctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
        {\quoteTy A} \equiv A}}
  \end{mathpar}
  \caption{Formation rules}
  \label{fig:formation}
\end{figure}

\steven{need to write the rules for the equalities of codes
  $M \equiv \quoteTy {\unquoteTy M} : U$ in an equational theory figure for
  non-linear stuff? Same for the codes in universe $L$.

  I will put the nonlinear equational rules, which are so
  unnecessary, in the appendix}
\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

    \inferrule{\nonlinterm \Gamma {M \equiv N} X}{\nonlinterm {\Gamma} {refl} {M
      =_{X} N}}

@@

    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}
  \end{mathpar}
  \caption{Non-linear Typing}
  \label{fig:non-linear-typing}
\end{figure}

\subsection{Non-linear Typing}
The non-linear types of \theoryabbv include the standard dependent
types for $\Pi,\Sigma$, extensional equality, natural numbers,
booleans, unit and empty types, and we assume function
extensionality\cite{Hofmann_1997}. We present a selection of
the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types} \steven{Do we all need non-linear terms? That seems
extra} of the appendix.
First, we include universe types $U$ of
small non-linear types and $L$ of linear types. These are defined as
universes ``ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
types are needed so that we can define types by recursion on natural
numbers. Next, we include a non-linear type $\ltonl A$ where $A$ is a
linear type. The intuition for this type is that its elements are the
linear terms that are ``resource free'': its introduction rule says we
can construct an $\ltonl A$ when we have a linear term of type $A$
with no free linear variables. This type is used extensively in our
examples, playing a similar role to the $!$ modality of ordinary
linear logic or the persistence modality $\square$ of separation logic
\cite{girard_linear_1987,jung_higher-order_2016}.

%% \begin{figure}
%%   \[
%%   \begin{array}{rrcl}
%%     \textrm{non-linear contexts} & \Gamma & ::= & \cdot \pipe \Gamma,x:X \\
%%     \textrm{non-linear types} & X , Y , Z & ::= & \mathsf{Nat} \pipe
%%                                                   \mathsf{Bool} \pipe
%%                                                   \mathsf{Empty} \pipe
%%                                                   \mathsf{Unit} \pipe
%%                                                   \mathsf{SPF}\,X\pipe U \pipe L
%%                                                   \pipe \ltonl A \pipe \sum\limits_{x:X} Y
%%                                                   \pipe \prod\limits_{x:X} Y\pipe M =_A
%%                                                   N
%%   \end{array}
%%   \]
%%   \caption{Non-linear Types}
%% \end{figure}
\begin{figure}
  \begin{footnotesize}
    \begin{mathpar}
    \inferrule{}{\Gamma \vdash U \isTy}\and
    \inferrule
    {\Gamma \vdash M : U}
    {\Gamma \vdash \unquoteTy M \isTy}\and
    \inferrule
    {\Gamma \vdash X \isTy}
    {\Gamma \vdash \quoteTy X : U}\and
    \unquoteTy{\quoteTy{X}} \equiv X \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}

    \inferrule{}{\Gamma \vdash L \isTy}\and
    \inferrule
    {\Gamma \vdash M : L}
    {\Gamma \vdash \unquoteTy M \isLinTy}\and
    \inferrule
    {\Gamma \vdash A \isLinTy}
    {\Gamma \vdash \quoteTy A : L}\and
    \unquoteTy{\quoteTy{A}} \equiv A \and \inferrule{\Gamma \vdash M : U}{\Gamma\vdash \quoteTy{\unquoteTy{M}}\equiv M : U}

    \inferrule
    {\Gamma \vdash A \isLinTy}
    {\Gamma \vdash \ltonl { A } \isTy}
    \and
    \inferrule{\Gamma ; \cdot \vdash e : A}{\Gamma \vdash e : \ltonl A}\and
    \inferrule{\Gamma \vdash M : \ltonl A}{\Gamma ; \cdot \vdash M : A} \and
  \end{mathpar}
  \end{footnotesize}
  \caption{Non-linear types (selection)}
  \label{fig:non-linear-types}
\end{figure}
\subsection{Linear Typing}

\steven{todo}
We give an overview of the linear types and terms in
\Cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and included in the
appendix. First, the linear variable rule says that a linear variable
can be used if it is the \emph{only} variable in the context.

@@

Because we are non-commutative, there are two function types: $A \lto
B$ and $B \tol A$, which have similar $\lambda$ introduction forms and
application elimination forms. The difference between these is that
the introduction rule for $A \lto B$ adds a variable to the right side
of the context, whereas the introduction rule\footnote{Rules for $\tol$ elided
  here, but found in the appendix in \cref{fig:extend-lin-terms}.} for $B \tol A$ adds a
variable to the left side of the context. In our experience, because
by convention parsing algorithms parse from left-to-right, we rarely
need to use the $B \tol A$ connective. As we have already seen, the
$\lto$ connective is frequently used in conjunction with the
$\uparrow$ connective so that we can abstract non-linearly over linear
functions.

@@

\begin{corollary}
  \label{cor:binary-mono}
  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms --- i.e. if
  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
\end{corollary}

\steven{cite reference for distributivity implies coprojections mono. Cockett
  ``Introduction to Distributive Categories''? Awodey?}

@@

So far, our calculus is a somewhat generic combination of dependent
types with non-commutative linear types. In order to carry out formal
grammar theory and define parsers, we need only add a few
grammar-specific constructions. First for each character in our
alphabet, we add a corresponding linear type $c$. We can then define a
non-linear type $\CharGram$ as the disjunction of all of these characters,
and define a type $\StringGram$ as the Kleene star of $\CharGram$, i.e. as an
inductive linear type.
Then we add a function $\mathsf{read} :
\ltonl{(\top\lto \StringGram)}$ that intuitively ``reads'' the input
string from the input and makes it available. It is important that the
input type of $\mathsf{read}$ is $\top$, which can control any amount
of resources, and not $\I$ which controls no resources. Further, we
add an axiom that $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$,
i.e., that if we have a string, but then throw it away and read
it from the input, then in fact that is equivalent to the string we
were given originally. This ensures that the elements of the $\StringGram$
type always stand for the actual input string in our reasoning. In the
next section, we will show how these basic principles are enough to
provide a basis for verified parsing and formal grammar theory.

@@

\steven{
With \cref{ax:dist}, it is straightforward to implement a parser for this lookahead automaton,
generalizing the approach for deterministic automata. Without access to
distributivity, we may define a rudimentary lookahead operation via the chain of
equivalences,

@@

When defining a parser, if the lookahead character is a right paren we would
like to apply $\texttt{lookAheadRP}$; otherwise, apply $\texttt{lookAheadNot}$.
However, without distributivity we have no means of relating the bit of
information learned by looking ahead to the input $A~n~b$ parse.
That is, to choose which constructor to apply we would need the end of this
chain of equivalences to be a binary sum rather than a binary product,
necessitating the distributivity axiom.
}

@@

The interpretation of dependent types as sets is standard
\cite{Hofmann_1997}. We present the concrete descriptions of the
semantics of linear types, as well as our non-standard non-linear
types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
single parse precisely when the input string consists of the single
character. The grammar for the unit similarly has a single parse for
the empty string.  A parse of the tensor product $A \otimes B$
consists of a \emph{splitting} of the empty string into a prefix $w_1$
and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
except the $B$ parse is for the reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
$\sem{g}$.
%
The universe $L$ of linear types is interpreted as the set of all
small grammars.
\begin{figure}
  \begin{footnotesize}
    \begin{align*}
    \sem{c}\gamma\,w &= \{ c | w = c\}\\
    \sem{I}\gamma\,w &= \{ () \pipe w = \varepsilon \}\\
    \sem{A \otimes B}\gamma\,w &= \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    \sem{A \lto B}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    \sem{B \tol A}\gamma\,w &= \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    \sem{\bigoplus_{x:X} A}\gamma\,w &= \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    \sem{\bigwith_{x:X} A}\gamma\,w &= \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w\\
    \sem{\uparrow A}\gamma &= \sem{A}\gamma\,\varepsilon\\
    \sem{\equalizer{a}{f}{g}}\gamma\,w &= \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    \sem{L}\gamma &= \Grammar_0\\
    \sem{\SPF\,X}\gamma &= \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    \sem{\el(F)}\gamma\,G &= \sem{F}\gamma\,G\\
    \sem{\map(F)}\gamma\,f &= \sem{F}\gamma\,f\\
    \sem{\mu A}\gamma &= \mu (\sem{A}\gamma)\\
  \end{align*}
  \end{footnotesize}
  \caption{Grammar Semantics}
  \label{fig:semantics}
\end{figure}

@@

The remaining details of the interpretation of linear terms as parse
transformers and verification of the equational axioms is a relatively
straightforward extension of existing semantics of linear logic in
monoidal categories, and is included in the appendix\cite{seely89}.

@@

The most similar prior work to our own is Luo's Lambek calculus with
dependent types \cite{luo}. Their design differs from our own in
allowing linear types to depend on other linear types and supporting
directed $\Pi$ and $\Sigma$ types that have no analog in our
system. They do not provide a semantics for these connectives, and it
is unclear how to interpret their connectives in our grammar
semantics. Further, we provide several examples showing that
\theoryabbv is a practical system for describing grammar formalisms
and parsers, and it is unclear if these could be implemented in their calculus.

%% \subsubsection{Typed Grammars}
Frisch and Cardelli introduced the use of a simple type system to reason about regular expressions up to weak equivalence \cite{frischCardelli}.
Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
Our design of \theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.

@@

We also aim to adapt the approach used for our context-free grammar parserss for
LL and LR/LALR with the aim of developing a shared library of intrinsically
verified parsing utilities that can be usefully shared across different
parsers. We would also like to investigate how high-performance verified parsers
can be implemented in \theoryabbv. This might be done by developing an efficient
compiler for \theoryabbv directly or by developing a verified compiler to an
imperative system within \theoryabbv itself.

@@

\begin{figure}
  \begin{mathpar}
    \inferrule{~}{\ctxwff \cdot}
    \and
    \inferrule{\ctxwff \Gamma \\ \ctxwffjdg \Gamma X}{\ctxwff {\Gamma, x : X}}
    \inferrule{\Gamma \vdash X : U_i}{\ctxwffjdg \Gamma X}

>>>>

\newcommand{\Subst}{\textrm{Subst}}
\newcommand{\SPF}{SPF}
\newcommand{\Var}{\mathsf{Var}}
\newcommand{\K}{\mathsf{K}}
\newcommand{\map}{\mathsf{map}}
\newcommand{\roll}{\mathsf{roll}}
\newcommand{\fold}{\mathsf{fold}}
\newcommand{\inl}{\mathsf{inl}}
\newcommand{\inr}{\mathsf{inr}}
\newcommand{\sem}[1]{\llbracket{#1}\rrbracket}
\newcommand{\semg}[1]{\sem{#1}\gamma}
\newcommand{\cat}[1]{\mathbf{#1}}
\newcommand{\lto}{\multimap}
\newcommand{\tol}{\mathrel{\rotatebox[origin=c]{180}{$\lto$}}}
\newcommand{\StringSem}{\textbf{String}}
\newcommand{\CharGram}{\texttt{Char}}
\newcommand{\StringGram}{\texttt{String}}
\newcommand{\Set}{\mathbf{Set}}
\newcommand{\Syn}{\mathbf{Synx}}
\newcommand{\SemAct}{\mathbf{SemAct}}
\newcommand{\Gr}{\mathbf{Gr}}
\newcommand{\Grammar}{\mathbf{Gr}}
\newcommand{\semcat}{\mathbf{C}}
\newcommand{\Type}{\mathbf{Type}}
\newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Bool}{\mathtt{Bool}}
\newcommand{\true}{\mathtt{true}}
\newcommand{\false}{\mathtt{false}}
\newcommand{\nat}{\mathbb{N}}
\newcommand{\theoryname}{Dependent Lambek Calculus\xspace}
\newcommand{\theoryabbv}{$\textrm{Lambek}^D$~}
\newcommand{\lnld}{$\textrm{LNL}_D$}

@@

\newcommand{\ctxwff}[1]{#1 \isCtx}
\newcommand{\ctxwffjdg}[2]{#1 \vdash #2 \isTy}
\newcommand{\linctxwff}[2]{#1 \vdash #2 \isLinCtx}
\newcommand{\linctxwffjdg}[2]{#1 \vdash #2 \isLinTy}
\newcommand{\nonlinterm}[3]{#1 \vdash #2 : #3}
\newcommand{\linterm}[4]{#1 ; #2 \vdash #3 : #4}
\newcommand{\nonlineq}[4]{#1 \vdash #2 \equiv #3 : #4}
\newcommand{\lineq}[5]{#1 ; #2 \vdash #3 \equiv #4 : #5}

\newcommand{\LLL}{\textrm{LL}}
\newcommand{\LRR}{\textrm{LR}}
\newcommand{\LALRR}{\textrm{LALR}}
\newcommand{\LL}[1]{\LLL(#1)}
\newcommand{\LR}[1]{\LRR(#1)}
\newcommand{\LALR}[1]{\LALRR(#1)}

@@

\section{Introduction}
\label{sec:intro}
Parsing structured data from untrusted input is a ubiquitous task in
computing. Any formally verified software system that interacts with
the outside world must contain some parsing component. For example, in
an extensive experiment finding bugs in C compilers
\cite{yangFindingUnderstandingBugs}, an early version of the formally
verified CompCert C compiler only contained bugs in the then
unverified parsing component \cite{leroy_formal_2009}. Bugs in parsers
undermine the overall correctness theorem for a verified system: an
incorrectly parsed C program will be compiled correctly but this is
not very useful if it did not correctly correspond to the actual
source program. Eventually, a correct parser was implemented using an
automaton that is formally verified to implement an $\LR{1}$ grammar
\cite{jourdanValidatingLRParsers2012}.

@@

Prior approaches to verified parsing focus on verification of a
particular grammar formalism such as non-left-recursive grammars or
$\LL{1}$ grammars
\cite{lasserCoStarVerifiedALL2021,EdelmannZippy2020,danielssonTotalParserCombinators2010}.
Each new grammar formalism is extended with its own independent
verified implementation.

@@

To show the feasibility of our design, we have implemented \theoryabbv
as a shallowly embedded domain-specific language in the Cubical Agda
proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
grammars and parsers in our system including regular expressions,
non-deterministic and deterministic automata, as well as some example
context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
automata. Throughout this paper, we will use \agdalogo~ to mark
results that are mechanized in our Agda development.

@@

The formation rules for the judgments of \theoryabbv are shown in
\Cref{fig:formation}. $\Gamma$ stands for non-linear contexts; $X,Y,Z$
stand for non-linear types; $M,N$ stand for non-linear terms, these
act as in an ordinary dependent type theory; $\Delta$ stands for
linear contexts; $A,B,C$ for linear types; and, $e,f,g$ for linear
terms. These contexts, types and terms are allowed to depend on an
ambient non-linear context $\Gamma$, but note that linear types $A$
cannot depend on any \emph{linear} variables in $\Delta$. We include
definitional equality judgments for both kinds of type and term
judgments as well. Additionally, we have a judgment $\Gamma \vdash X
\isSmall$ which is used in the definition of universe types.

@@

    \inferrule{\linctxwff \Gamma \Delta \and \linctxwffjdg \Gamma A}{\linctxwff
      {\Gamma} {\Delta , a : A}}
  \end{mathpar}
  \caption{Context well-formedness rules}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \footnotesize
    \boxed{\ctxwffjdg \Gamma X}

@@

    \inferrule{\ctxwffjdg \Gamma X \and \nonlinterm \Gamma M X \and \nonlinterm \Gamma N X}{\ctxwffjdg \Gamma {M =_{X} N}}

@@

    \inferrule{\linctxwffjdg \Gamma X}{\linctxwffjdg \Gamma {\unquoteTy
        {\quoteTy X} \equiv X}}

@@

    \inferrule{\linctxwffjdg \Gamma A}{\linctxwffjdg \Gamma {\unquoteTy
        {\quoteTy A} \equiv A}}
  \end{mathpar}
  \caption{Type formation rules}
  \label{fig:formation}
\end{figure}

\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

@@

    \inferrule{\linterm \Gamma \cdot e A}{\nonlinterm \Gamma e \ltonl{A}}

    \inferrule{\nonlinterm \Gamma {M \equiv N} {X}}
      {\nonlinterm \Gamma {refl} {M =_{X} N}}
  \end{mathpar}
  \caption{Non-linear typing (selection)}
  \label{fig:non-linear-types}
\end{figure}

\subsection{Non-linear Typing}
The non-linear types of \theoryabbv include the standard dependent
types for $\Pi$,$\Sigma$, extensional equality, natural numbers,
\steven{,just take Booleans to be implemented via $1$ and $\Sigma$,} unit, and empty types, and we assume function
extensionality\cite{Hofmann_1997}. We present a selection of
the non-linear type constructions in
\Cref{fig:non-linear-types} and provide the rest in
\cref{fig:full-non-linear-types, fig:jdg-eq-nonlinear} of the appendix.
First, we include universe types $U$ of
small non-linear types and $L$ of linear types. These are defined as
universes ``ala Coquand'' in that we define a judgment saying when a
non-linear type is small and define the universe to internalize
precisely this judgment \cite{coquandPresheafModel,lmcs:7713}. These universe
types are needed so that we can define types by recursion on natural
numbers. As in \cite{krishnaswami_integrating_2015}, the extensional equality type
$M =_{X} N$ used here has the introduction form $refl$, but no
elimination due to the presence of equality reflection. Next, we include a non-linear type $\ltonl A$ where $A$ is a
linear type. The intuition for this type is that its elements are the
linear terms that are ``resource free'': its introduction rule says we
can construct an $\ltonl A$ when we have a linear term of type $A$
with no free linear variables. This type is used extensively in our
examples, playing a similar role to the $!$ modality of ordinary
linear logic or the persistence modality $\square$ of separation logic
\cite{girard_linear_1987,jung_higher-order_2016}.

\subsection{Linear Typing}
We give the rules for linear type
formation in \cref{fig:formation} and the definition of linear terms in
\cref{fig:linear-terms}. The equational theory for these
types is straightforward $\beta\eta$ equivalence and included in the
appendix. First, the linear variable rule says that a linear variable
can be used if it is the \emph{only} variable in the context.

@@

Because we are non-commutative, there are two function types: $A \lto
B$ and $B \tol A$, which have similar $\lambda$ introduction forms and
application elimination forms. The difference between these is that
the introduction rule for $A \lto B$ adds a variable to the right side
of the context, whereas the introduction rule for $B \tol A$ adds a
variable to the left side of the context. In our experience, because
by convention parsing algorithms parse from left-to-right, we rarely
need to use the $B \tol A$ connective. As we have already seen, the
$\lto$ connective is frequently used in conjunction with the
$\uparrow$ connective so that we can abstract non-linearly over linear
functions.

@@

\steven{I'm not sure if this is the proper way to cite the corollary, especially
because the citation is a reference, not the originator of the idea}
\begin{corollary}[\cite{Cockett_1993}]
  \label{cor:binary-mono}
  Distributivity implies that the coprojections $\inl : A \to A \oplus B$,
  $\inr : B \to A \oplus B$ of a binary sum are monomorphisms --- i.e. if
  $\inl\,a \equiv \inl\,a'$, then $a \equiv a'$.
\end{corollary}

@@

So far, our calculus is a somewhat generic combination of dependent
types with non-commutative linear types. In order to carry out formal
grammar theory and define parsers, we need only add a few
grammar-specific constructions. First for each character in our
alphabet, we add a corresponding linear type $c$. We can then define a
non-linear type $\CharGram$ as the disjunction of all of these characters,
and define a type $\StringGram$ as the Kleene star of $\CharGram$, i.e. as an
inductive linear type.
Then we add a function $\mathsf{read} :
\ltonl{(\top\lto \StringGram)}$ that intuitively ``reads'' the input
string from the input and makes it available. It is important that the
input type of $\mathsf{read}$ is $\top$, which can control any amount
of resources, and not $\I$ which controls no resources.

\begin{axiom}
  \label{ax:string-top}
   $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$. That is, $\StringGram$ is strongly equivalent to $\top$.
\end{axiom}

If we have a string, but then throw it away and read
it from the input, then we, in fact, recover the original string.
This ensures that the elements of the $\StringGram$
type always stand for the actual input string in our reasoning. In the
next section, we will show how these basic principles are enough to
provide a basis for verified parsing and formal grammar theory.

@@

With \cref{ax:dist}, it is straightforward to implement a parser for this lookahead automaton,
generalizing the approach for deterministic automata. Without access to
distributivity, we may define a rudimentary lookahead operation via the chain of
equivalences,

@@

When defining a parser, if the lookahead character is a right paren we would
like to apply $\texttt{lookAheadRP}$; otherwise, apply $\texttt{lookAheadNot}$.
However, without distributivity we have no means of relating the bit of
information learned by looking ahead to the input $A~n~b$ parse.
That is, to choose which constructor to apply we would need the end of this
chain of equivalences to be a binary sum rather than a binary product,
necessitating the distributivity axiom.

@@

The interpretation of dependent types as sets is standard
\cite{Hofmann_1997}. We present the concrete descriptions of the
semantics of linear types, as well as our non-standard non-linear
types in \Cref{fig:semantics}. The grammar for a literal $c$ has a
single parse precisely when the input string consists of the single
character. The grammar for the unit similarly has a single parse for
the empty string.  A parse of the tensor product $A \otimes B$
consists of a \emph{splitting} of the empty string into a prefix $w_1$
and suffix $w_2$ along with an $A$ parse of $w_1$ and $B$ parse of
$w_2$. A parse of $\bigoplus_{x:X} A$ is a pair of an element of the
set $X$ and a parse of $A(x)$, while dually a parse of $\bigwith_{x:X}
A$ is a \emph{function} taking any $x:X$ to a parse of $A(x)$. A
$w$-parse of $A \lto B$ is a function that takes an $A$ parse of some
other string $w'$ to a $B$ parse of $ww'$, and $B \tol A$ is the same
except the $B$ parse is for the reversed concatenation $w'w$.
%
The set $\uparrow A$ is the set of parse for the empty string for
$A$. This definition means that $\sem{\uparrow (A \lto B)}$ (or $\sem{\uparrow (B \tol A)}$) is
equivalent to the set of parse transformers:
\[ \sem{\uparrow (A \lto B)}\gamma = \sem{A \lto B}\gamma\,\varepsilon = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'\]
Next, a parse in the equalizer $\equalizer{a}{f}{g}$ is defined as
a parse in $\sem{A}$ that is mapped to the same parse by the parse transformers $\sem{f}$ and
$\sem{g}$.
%
The universe $L$ of linear types is interpreted as the set of all
small grammars.
\begin{figure}
  \begin{minipage}[t]{.5\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{c}\gamma\,w = \{ c | w = c\}\\
    &\sem{I}\gamma\,w = \{ () \pipe w = \varepsilon \}\\
    &\sem{A \otimes B}\gamma\,w = \{ (w_1,w_2,a,b) \pipe w_1w_2 = w \wedge a \in \sem{A}\gamma\,w_1 \wedge b \in \sem{B}\gamma\,w_2 \}\\
    &\sem{A \lto B}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,ww'\\
    &\sem{B \tol A}\gamma\,w = \prod_{w'} \sem{A}\gamma\,w' \to \sem{B}\gamma\,w'w\\
    &\sem{\bigoplus_{x:X} A}\gamma\,w = \{ (x, a) \pipe x \in \sem{X}\gamma \wedge a \in \sem{A}(\gamma,x)w \}\\
    &\sem{\bigwith_{x:X} A}\gamma\,w = \prod_{x\in \sem{X}\gamma} \sem{A}(\gamma, x)\,w
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}%
  \begin{minipage}[t]{.5\textwidth}
  \begin{footnotesize}
    \begin{flalign*}
    &\sem{\uparrow A}\gamma = \sem{A}\gamma\,\varepsilon\\
    &\sem{\equalizer{a}{f}{g}}\gamma\,w = \{ a \in \sem{A}\gamma\,w \pipe \sem{f}\gamma\,w\,a = \sem{g}\gamma\,w\,a \} \\
    &\sem{L}\gamma = \Grammar_0\\
    &\sem{SPF\,X}\gamma = \textrm{DepPolyFunctor}(\sem{X}\gamma\times \Sigma^*,\Sigma^*)\\
    &\sem{\el(F)}\gamma\,G = \sem{F}\gamma\,G\\
    &\sem{\map(F)}\gamma\,f = \sem{F}\gamma\,f\\
    &\sem{\mu A}\gamma = \mu (\sem{A}\gamma)
  \end{flalign*}
  \end{footnotesize}
  \end{minipage}
  \caption{Grammar Semantics}
  \label{fig:semantics}
\end{figure}

@@

The remaining details of the interpretation of linear terms as parse
transformers and verification of the equational axioms --- as well as
\cref{ax:dist,ax:disjointness,ax:string-top} --- is a relatively
straightforward extension of existing semantics of linear logic in
monoidal categories, and is included in the appendix\cite{seely89}.

@@

The most similar prior work to our own is Luo's Lambek calculus with
dependent types \cite{luo}. They present two systems: one like ours where linear
types may only depend on non-linear ones, and another that
allows linear types to depend on other linear types and supports
directed $\Pi$ and $\Sigma$ types that have no analog in our
system. They do not provide a semantics for these connectives, and it
is unclear how to interpret their connectives in our grammar
semantics. Further, we provide several examples showing that
\theoryabbv is a practical system for describing grammar formalisms
and parsers, and it is unclear if these could be implemented in their calculus.

%% \subsubsection{Typed Grammars}
Frisch and Cardelli introduced the use of a simple type system to reason about
regular expressions up to weak equivalence \cite{frischCardelli}. \steven{They
  use their system in the design of a linear-time algorithm for greedy regular
  expression matching. It is yet unclear if greediness can be encoded in
  \theoryname; however, it appears that our approach in determinizing an NFA is
  quite similar to their algorithm. The biggest difference being that our
  approach uses a global disambiguation strategy while theirs is a local, greedy strategy.
}
Henglein and Nielsen build on this type-theoretic view of grammars, pointing out that the values of the types correspond precisely to the parse trees \cite{henglein_regular_2011}.
Our design of \theoryabbv extends this view of grammars as types to a much broader class of grammars, as well as providing a syntax and equational theory for \emph{parse transformers}, showing that the terms can be viewed not just as parse trees, but as intrinsically verified parsers.

@@

We also aim to adapt the approach used for our context-free grammar parserss for
$\LLL$ and $\LRR$/$\LALRR$ with the aim of developing a shared library of intrinsically
verified parsing utilities that can be usefully shared across different
parsers. We would also like to investigate how high-performance verified parsers
can be implemented in \theoryabbv. This might be done by developing an efficient
compiler for \theoryabbv directly or by developing a verified compiler to an
imperative system within \theoryabbv itself.

@@

\begin{figure}
  \begin{mathpar}
  \footnotesize
    \boxed{\nonlinterm \Gamma M X}

    \inferrule
    {\nonlinterm \Gamma M \bot \and \ctxwffjdg \Gamma X}
    {\nonlinterm \Gamma {()} {\bot}}

    \inferrule{~}{\nonlinterm {\Gamma} {0} {Nat}}
